Informing policy via dynamic models: Eliminating cholera in Haiti

### 2022-09-12 ms.Rnw
Compiled merge
minor edits

@@

The four independent teams were given the task of estimating the
potential effect of prospective oral cholera vaccine (OCV) programs.
OCV is accepted as a safe and effective tool for controlling the spread of cholera, however the available stockpile and production of OCV doses is insufficient to meet global needs \citep{lee20, pezzoli20} \eic{NEW CHOLERA VACCINES, NEW STUDIES OF EXISTING VACCINES, AND INCREASED PRODUCTION VOLUMES ALL AROSE DURING 2010-2019. MAYBE SOMETHING LIKE:
Advances in OCV technology and vaccine availability raised the possibility of planning a national vaccination program \citep{lee20, pezzoli20}}
In the study, certain data were shared between the groups, including demography and vaccination history; vaccine efficacy was also fixed at a shared value between groups.
Beyond this, the groups made autonomous decisions on what to include and exclude from their models; this autonomy reduced the possible effect that assumptions about the dynamic system may have on the final conclusion of the study.
Despite this autonomy, and largely adhering to existing guidelines on creating models to inform policy \citep{behrend20,saltelli20}, the consensus across the four models was that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
This conclusion is inconsistent with the fact that there have been no confirmed cases since February, 2019 without additional vaccination efforts \citep{ferguson22}.

@@

\begin{tabular}{|c|c|c|c|c|}
\hline
 & \thead{Model~1} & \thead{Model~2} & \thead{Model~3 \\ (panelPOMP)} & \thead{Model~3 \\ (SpatPOMP)}
\\
\hline
\hline
\multirow{2}{*}{Log-likelihood} &
  $\Sexpr{myround(mod1_ll, 1)}$ &
  $\Sexpr{myround(mod2_ll, 1)}$ &
  \multirow{2}{*}{$\Sexpr{myround(mod3_ll[1], 1)}$\footnotemark[2]} &
  $\Sexpr{myround(mod3_spatPomp_res$ll, 1)}$ \\
    & ($\Sexpr{myround(mod1_lee_ll, 1)}$)\footnotemark[1] &
  ($\Sexpr{myround(h2_fit$leeFit_ll, 1)}$) &
  & ($\Sexpr{myround(max(lee3_results[["all_ll_evals"]]), 1)}$)\footnotemark[3]
\\
\hline
Number of &
  $\Sexpr{as.character(mod1_n_params)}$ & $\Sexpr{as.character(mod2_n_params)}$ & $\Sexpr{as.character(mod3_n_params)}$ & $\Sexpr{mod3_spatPomp_n_params}$ \\
 Fit Parameters & (\Sexpr{as.character(mod1_lee_n_params)}) & ($\Sexpr{as.character(h2_fit$leeFit_n_params)}$) & (29) & ($\Sexpr{as.character(lee3_results$n_params, 1)}$)
\\
\hline
\multirow{2}{*}{AIC} &
  $\Sexpr{myround(mod1_aic, 1)}$ & $\Sexpr{myround(mod2_aic, 1)}$ & \multirow{2}{*}{$\Sexpr{myround(mod3_aic, 1)}$\footnotemark[2]} & $\Sexpr{myround(mod3_spatPomp_aic, 1)}$ \\
  & ($\Sexpr{myround(mod1_lee_aic, 1)}$)\footnotemark[1] & ($\Sexpr{myround(mod2_lee_aic, 1)}$) &  & ($\Sexpr{myround(mod3_lee_coupled_aic, 1)}$)\footnotemark[3]
\\
\hline
\thead{Log-ARMA(2,1) \\ Log-likelihood} &
  $\Sexpr{myround(ARMA_benchmarks$m1, 1)}$ &
  $\Sexpr{myround(ARMA_benchmarks$m2, 1)}$ &
  $\Sexpr{myround(ARMA_benchmarks$m3, 1)}$ &
  $\Sexpr{myround(ARMA_benchmarks$m3, 1)}$
\\
\hline
\end{tabular}
\caption{\label{tab:likes}Log-likelihood values for each models compared to their ARMA benchmarks.
Values in parenthesis are corresponding values using \citet{lee20} parameter estimates.
\eic{I THINK THAT THE NOTES SHOULD GO IN THE CAPTION RATHER THAN BEING FOOTNOTES.}
\eic{WE SHOULD REVIEW THIS TABLE IN A MEETING. RIGHT NOW, I'M SURPRISED BY A FEW THINGS. WHY IS THE LEE ET AL MODEL 2 LIKELIHOOD HIGHER THAN OURS? IS THE (29) IN THE PANELPOMP COLUMN A TYPO, SINCE WE DON'T COMPARE TO LEE ET AL IN THIS COLUMN? NESTING RELATIONSHIP FOR THE SPATPOMP VERSION OF MODEL 3 MAKES IT HARD TO EXPLAIN THE RATHER LARGE DECREASE IN LOGLIK. AT FACE VALUE, THIS SEEMS TO MEAN THAT THE COUPLING IS SUFFICIENTLY POORLY SPECIFIED THAT WE ARE BETTER OFF WITHOUT IT. I SEEM TO RECALL THAT OUR UNDERSTANDING INVOLVES THE WAY PANELPOMP USES A COVARIATE THAT DEPENDS ON CASES IN OTHER DEPARTMENTS. SINCE LEE ET AL ALSO DO THAT, COULD IT BE THAT THEIR ANALYSIS IS MORE CLOSELY RELATED TO THE PANELPOMP VERSION THAN THE SPATPOMP VERSION?}
}
\end{table}

@@

\eic{I THINK WE WILL HAVE TO EXPLAIN THIS CAREFULLY IF THE POINT IS TO APPEAR NONTRIVIAL. IT IS WORTH NOTING THAT LEE ET AL DID NOT PROPERLY FORECAST FROM THE BEST ESTIMATE OF THE CURRENT STATE. HOWEVER, IT WILL BE OBVIOUS TO MOST SATISTICAL READERS THAT THIS IS NOT THE BEST THING TO DO. IT MAY NOT BE SO OBVIOUS THAT THE DIFFERENCE IS IMPORTANT IF AND ONLY IF A DETERMINISTIC MODEL IS INADEQUATE. THE QUESTION OF WHETHER TO USE A DETERMINISTIC MODEL REMAINS OF CURRENT INTEREST...}
As an example, we note that simulations from a well-fit mechanistic model may closely resemble the observed data $y_{1:N}^*$.
In such comparisons, these simulations are random draws from the complete estimated joint distribution $f_{\bm{X}^{(\modelCounter)}_{0:N}, \bm{Y}^{(\modelCounter)}_{1:N}}\left(x_{0:N}, y_{1:N}\mid \hat{\paramVec}\right)$, where $\modelCounter$ indexes the model used for simulations.
It can therefore be tempting to use simulations from this model up to time $N+s$, $f_{\bm{X}^{(\modelCounter)}_{0:N+s}, \bm{Y}^{(\modelCounter)}_{1:N+s}}\left(x_{0:N+s}, y_{1:N+s}\mid \hat{\paramVec}\right)$, with $s\geq 0$ to project the dynamic system up to a future time $N+s$, as it has been done in previous studies \citep{lee20,???}.
This approach, however, does not take advantage of the information about the state of the system that is contained in the observed data.
Note that each of the models in question (and, more generally, all models that are described as POMPs) are Markovian, that is, the history of the process $\{\bm{X}^{(\modelCounter)}_s, s\leq t\}$ for model $\modelCounter$ is uninformative about the future of the process $\{\bm{X}^{(\modelCounter)}_s, s\geq t\}$, given the current state $\bm{X}^{(\modelCounter)}_t$.
In other words, observing the data $\bm{Y}^{(\modelCounter)}_N = y_N^*$ at time $N$ provides more information about the future state of the system than the initial conditions.
In this case, draws from the conditional density $f_{\bm{X}^{(\modelCounter)}_{N:s}, \bm{Y}^{\modelCounter}_{N:s}}\big(x_{N:s}, y_{N:s}\mid \hat{\paramVec}, \bm{X}^{(\modelCounter)}_{N} = \bm{x}^{(\modelCounter)}_N\big)$ should be preferred as forecasts, as these simulations account for the most recent known state of the system.

$\bm{X}^{(\modelCounter)}_N$, however, is unobservable and therefore draws from the desired conditional density are unobtainable.
Informed estimates of $\bm{X}^{(\modelCounter)}_N$ given the observed data $\bm{Y}^{(\modelCounter)}_{1:N} = y_{1:N}^*$ can easily be obtained, however, via the filtering distribution.
Let $\hat{\bm{X}}^{(\modelCounter), i}_N, \, i \in {1, 2, \ldots, J}$ be $\iid$ draws from the filtering distribution at time $N$, with density $\hat{\bm{X}}^{(\modelCounter), i}_N \sim f_{\bm{X}^{(\modelCounter)}_N|\bm{Y}^{(\modelCounter)}_{1:N}}(x_N \mid \hat{\paramVec}, y_{1:N}^*)$.
A single model forecast can then be obtained by simulating from the model $f_{\bm{X}^{(\modelCounter)}_{N:s}, \bm{Y}^{(\modelCounter)}_{N:s}}\big(x_{N:s}, y_{N:s}\mid \hat{\paramVec}, \bm{X}^{(\modelCounter)}_{N} = \hat{\bm{X}}^{(\modelCounter), i}_N\big)$.
Intuitively, simulating the model starting at the filtering distribution of the most recently available time point is a more appropriate way to project a stochastic dynamic system into the future, as it is not expected that each simulation from initial conditions will result in a latent state at time $N$ consistent with the model and the observation $\bm{Y}^{(\modelCounter)}_N = y_N^*$.
In this particular case study, projecting the future state of the cholera epidemic in Haiti starting from the draws from the filtering distribution allows the proposed models to benefit from the fact that very few cholera cases had been observed between 2018 and 2019, and that cases appear to be decreasing (i.e., the number of susceptible individuals may be small).

@@

In a similar fashion, one can take advantage of certain mechanistic features contained in a particular model in order to make inference on a system.
Examples of this are as diverse as estimating the effective reproductive number ($R_0$) of an infectious disease \citep{he10} or investigating interactions between pedestrians and autonomous vehicles \citep{domeyer22}.
In our analysis, we demonstrate this ability by examining the results of fitting the flexible cubic spline term in Model~1 (Eq.~\myeqref{model1:lambda}--\myeqref{model1:beta}), which allows for a flexible estimation of seasonality in the force of infection.
After fitting the model, we explore potential patterns in the seasonal transmission rate by plotting the average value of $\transmission$ in a typical year.
Fig.~\ref{fig:h1SeasRain} shows that the estimated seasonal transmission rate $\transmission$ mimics the rainfall dynamics in Haiti, despite Model~1 not having access to rainfall data.
This result provides evidence that rainfall is a potential driver of cholera infections in Haiti.

@@

Simulations from probabilistic models (Models~1 and~3) represent possible trajectories of the dynamic system under the scientific assumptions of the models.
In this case study, estimates of the probability of cholera elimination can therefore be obtained as the proportion of simulations from the fitted model that result in cholera elimination.
The results of these projections are summarized in Figs.~\ref{fig:Mod1Scenarios}--\ref{fig:elimProbs}.
These results suggest that cholera elimination was likely, even without increased vaccination efforts, which is consistent with observed reality \citep{ferguson22}. \eic{I LIKE THE IDEA OF REFERENCING POST-EPIDEMIC REFLECTIONS. UNLESS THERE ARE NO ALTERNATIVES, IT IS BEST TO AVOID CITING A NON-PEER-REVIEWED ARTICLE SUCH AS THIS. HOW ABOUT \citep{piarroux22} OR \citep{trevisin22}?}

@@

\begin{table}
\small

\hspace{-10mm}\begin{tabular}{|l@{}|l@{}c@{}|l@{}c@{}|l@{}c@{}|}

@@

>>>>

\newcolumntype{t}{>{\tiny}c}

@@

The four independent teams were given the task of estimating the
potential effect of prospective oral cholera vaccine (OCV) programs.
While OCV is accepted as a safe and effective tool for controlling the spread of cholera, the global stockpile of OCV doses remains limited \citep{pezzoli20}.
Advances in OCV technology and vaccine availability, however, raised the possibility of planning a national vaccination program \citep{lee20}.
In the study, certain data were shared between the groups, including demography and vaccination history; vaccine efficacy was also fixed at a shared value between groups.
Beyond this, the groups made autonomous decisions on what to include and exclude from their models; this autonomy reduced the possible effect that assumptions about the dynamic system may have on the final conclusion of the study.
Despite this autonomy, and largely adhering to existing guidelines on creating models to inform policy \citep{behrend20,saltelli20}, the consensus across the four models was that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
This conclusion is inconsistent with the fact that there have been no confirmed cases since February, 2019 without additional vaccination efforts \citep{ferguson22}.

@@

\begin{tabular}{|c|c|c|c|c|}
\hline
 & \thead{Model~1} & \thead{Model~2} & \thead{Model~3 \\ (panelPOMP)} & \thead{Model~3 \\ (SpatPOMP)}
\\
\hline
\hline
\multirow{2}{*}{Log-likelihood} &
  $\Sexpr{myround(mod1_ll, 1)}$ &
  $\Sexpr{myround(mod2_ll, 1)}$ &
  \multirow{2}{*}{$\Sexpr{myround(mod3_ll[1], 1)}$\footnotemark[2]} &
  $\Sexpr{myround(mod3_spatPomp_res$ll, 1)}$ \\
    & ($\Sexpr{myround(mod1_lee_ll, 1)}$)\footnotemark[1] &
  ($\Sexpr{myround(h2_fit$leeFit_ll, 1)}$) &
  & ($\Sexpr{myround(max(lee3_results[["all_ll_evals"]]), 1)}$)\footnotemark[3]
\\
\hline
Number of &
  $\Sexpr{as.character(mod1_n_params)}$ & $\Sexpr{as.character(mod2_n_params)}$ & \multirow{2}{*}{$\Sexpr{as.character(mod3_n_params)}$} & $\Sexpr{mod3_spatPomp_n_params}$ \\
 Fit Parameters & (\Sexpr{as.character(mod1_lee_n_params)}) & ($\Sexpr{as.character(h2_fit$leeFit_n_params)}$) && ($\Sexpr{as.character(lee3_results$n_params, 1)}$)
\\
\hline
\multirow{2}{*}{AIC} &
  $\Sexpr{myround(mod1_aic, 1)}$ & $\Sexpr{myround(mod2_aic, 1)}$ & \multirow{2}{*}{$\Sexpr{myround(mod3_aic, 1)}$\footnotemark[2]} & $\Sexpr{myround(mod3_spatPomp_aic, 1)}$ \\
  & ($\Sexpr{myround(mod1_lee_aic, 1)}$)\footnotemark[1] & ($\Sexpr{myround(mod2_lee_aic, 1)}$) &  & ($\Sexpr{myround(mod3_lee_coupled_aic, 1)}$)\footnotemark[3]
\\
\hline
\thead{Log-ARMA(2,1) \\ Log-likelihood} &
  $\Sexpr{myround(ARMA_benchmarks$m1, 1)}$ &
  $\Sexpr{myround(ARMA_benchmarks$m2, 1)}$ &
  $\Sexpr{myround(ARMA_benchmarks$m3, 1)}$ &
  $\Sexpr{myround(ARMA_benchmarks$m3, 1)}$
\\
\hline
\end{tabular}
\caption{\label{tab:likes}Log-likelihood values for each models compared to their ARMA benchmarks.
Values in parenthesis are corresponding values using \citet{lee20} parameter estimates.
\eic{I THINK THAT THE NOTES SHOULD GO IN THE CAPTION RATHER THAN BEING FOOTNOTES.}
\eic{WE SHOULD REVIEW THIS TABLE IN A MEETING. RIGHT NOW, I'M SURPRISED BY A FEW THINGS. WHY IS THE LEE ET AL MODEL 2 LIKELIHOOD HIGHER THAN OURS? IS THE (29) IN THE PANELPOMP COLUMN A TYPO, SINCE WE DON'T COMPARE TO LEE ET AL IN THIS COLUMN? NESTING RELATIONSHIP FOR THE SPATPOMP VERSION OF MODEL 3 MAKES IT HARD TO EXPLAIN THE RATHER LARGE DECREASE IN LOGLIK. AT FACE VALUE, THIS SEEMS TO MEAN THAT THE COUPLING IS SUFFICIENTLY POORLY SPECIFIED THAT WE ARE BETTER OFF WITHOUT IT. I SEEM TO RECALL THAT OUR UNDERSTANDING INVOLVES THE WAY PANELPOMP USES A COVARIATE THAT DEPENDS ON CASES IN OTHER DEPARTMENTS. SINCE LEE ET AL ALSO DO THAT, COULD IT BE THAT THEIR ANALYSIS IS MORE CLOSELY RELATED TO THE PANELPOMP VERSION THAN THE SPATPOMP VERSION?}
}
\end{table}

@@

% \eic{I THINK WE WILL HAVE TO EXPLAIN THIS CAREFULLY IF THE POINT IS TO APPEAR NONTRIVIAL. IT IS WORTH NOTING THAT LEE ET AL DID NOT PROPERLY FORECAST FROM THE BEST ESTIMATE OF THE CURRENT STATE. HOWEVER, IT WILL BE OBVIOUS TO MOST SATISTICAL READERS THAT THIS IS NOT THE BEST THING TO DO. IT MAY NOT BE SO OBVIOUS THAT THE DIFFERENCE IS IMPORTANT IF AND ONLY IF A DETERMINISTIC MODEL IS INADEQUATE.}

% As an example, we note that simulations from a well-fit mechanistic model may closely resemble the observed data $y_{1:N}^*$.
% In such comparisons, these simulations are random draws from the complete estimated joint distribution $f_{\bm{X}^{(\modelCounter)}_{0:N}, \bm{Y}^{(\modelCounter)}_{1:N}}\left(x_{0:N}, y_{1:N}\mid \hat{\paramVec}\right)$, where $\modelCounter$ indexes the model used for simulations.
% It can therefore be tempting to use simulations from this model up to time $N+s$, $f_{\bm{X}^{(\modelCounter)}_{0:N+s}, \bm{Y}^{(\modelCounter)}_{1:N+s}}\left(x_{0:N+s}, y_{1:N+s}\mid \hat{\paramVec}\right)$, with $s\geq 0$ to project the dynamic system up to a future time $N+s$, as it has been done in previous studies \citep{lee20,???}.
% This approach, however, does not take advantage of the information about the state of the system that is contained in the observed data.
% Note that each of the models in question (and, more generally, all models that are described as POMPs) are Markovian, that is, the history of the process $\{\bm{X}^{(\modelCounter)}_s, s\leq t\}$ for model $\modelCounter$ is uninformative about the future of the process $\{\bm{X}^{(\modelCounter)}_s, s\geq t\}$, given the current state $\bm{X}^{(\modelCounter)}_t$.
% In other words, observing the data $\bm{Y}^{(\modelCounter)}_N = y_N^*$ at time $N$ provides more information about the future state of the system than the initial conditions.
% In this case, draws from the conditional density $f_{\bm{X}^{(\modelCounter)}_{N:s}, \bm{Y}^{\modelCounter}_{N:s}}\big(x_{N:s}, y_{N:s}\mid \hat{\paramVec}, \bm{X}^{(\modelCounter)}_{N} = \bm{x}^{(\modelCounter)}_N\big)$ should be preferred as forecasts, as these simulations account for the most recent known state of the system.
The most recently available information about a dynamic system should be used to obtain reliable forecasts.
Using state space model $m$, this means that in order to obtain a forecast up to time $N+s$, where $s\geq 0$ and $N$ is the index for the last available data, we should simulate from the conditional density $f_{\bm{X}^{(\modelCounter)}_{N:s}, \bm{Y}^{\modelCounter}_{N:s}}\big(x_{N:s}, y_{N:s}\mid \hat{\paramVec}, \bm{X}^{(\modelCounter)}_{N} = \bm{x}^{(\modelCounter)}_N\big)$.
We note, however, that $\bm{X}^{(\modelCounter)}_N$ is unobservable and therefore draws from the desired conditional density are unobtainable.
Informed estimates of $\bm{X}^{(\modelCounter)}_N$ given the observed data $\bm{Y}^{(\modelCounter)}_{1:N} = y_{1:N}^*$ can easily be obtained using the filtering distribution.
Let $\hat{\bm{X}}^{(\modelCounter), i}_N, \, i \in {1, 2, \ldots, J}$ be $\iid$ draws from the filtering distribution at time $N$, with density $\hat{\bm{X}}^{(\modelCounter), i}_N \sim f_{\bm{X}^{(\modelCounter)}_N|\bm{Y}^{(\modelCounter)}_{1:N}}(x_N \mid \hat{\paramVec}, y_{1:N}^*)$.
A single model forecast can then be obtained by simulating from the model $f_{\bm{X}^{(\modelCounter)}_{N:s}, \bm{Y}^{(\modelCounter)}_{N:s}}\big(x_{N:s}, y_{N:s}\mid \hat{\paramVec}, \bm{X}^{(\modelCounter)}_{N} = \hat{\bm{X}}^{(\modelCounter), i}_N\big)$.
Obtaining forecasts in this fashion allows us to get results that are consistent with the model assumptions and the observed data.
% Intuitively, simulating the model starting at the filtering distribution of the most recently available time point is a more appropriate way to project a stochastic dynamic system into the future, as it is not expected that each simulation from initial conditions will result in a latent state at time $N$ consistent with the model and the observation $\bm{Y}^{(\modelCounter)}_N = y_N^*$.
In this particular case study, projecting the future state of the cholera epidemic in Haiti starting from the draws from the filtering distribution allows the proposed models to benefit from the fact that very few cholera cases had been observed between 2018 and 2019, and that cases appear to be decreasing (i.e., the number of susceptible individuals may be small).
This alone partially explains the unrealistic forecasts of \cite{lee20}, which were obtained by simulating the model from initial conditions;
Table~S7 of their supplement material shows that of their simulations that were consistent with observing zero cases in 2019, all forecasts resulted in the elimination of cholera.
\eic{THE QUESTION OF WHETHER TO USE A DETERMINISTIC MODEL REMAINS OF CURRENT INTEREST...}

@@

In a similar fashion, one can take advantage of certain mechanistic features contained in a particular model in order to make inference on a system.
Examples of this are as diverse as estimating the effective reproductive number ($R_0$) of an infectious disease \citep{he10} or investigating interactions between pedestrians and autonomous vehicles \citep{domeyer22}.
In our analysis, we demonstrate this ability by examining the results of fitting the flexible cubic spline term in Model~1 (Eq.~\myeqref{model1:lambda}--\myeqref{model1:beta}), which allows for a flexible estimation of seasonality in the force of infection.
After fitting the model, we explore potential patterns in the seasonal transmission rate by plotting the average value of $\transmission$ in a typical year.
Fig.~\ref{fig:h1SeasRain} shows that the estimated seasonal transmission rate $\transmission$ mimics the rainfall dynamics in Haiti, despite Model~1 not having access to rainfall data.
This result provides evidence that rainfall plays an important role in cholera transmission in Haiti, which is consistent with previous studies \citep{eisenberg13,rinaldo12}.
We also note the increase in the transmission rate during the Fall, which was also noticed at an earlier stage of the epidemic \citep{rinaldo12}.

@@

Simulations from probabilistic models (Models~1 and~3) represent possible trajectories of the dynamic system under the scientific assumptions of the models.
In this case study, estimates of the probability of cholera elimination can therefore be obtained as the proportion of simulations from the fitted model that result in cholera elimination.
The results of these projections are summarized in Figs.~\ref{fig:Mod1Scenarios}--\ref{fig:elimProbs}.
These results suggest that cholera elimination was likely, even without increased vaccination efforts, which is consistent with observed reality \citep{trevisin22}.

@@

\begin{table}
% \small
\resizebox{\textwidth}{!}{
\hspace{-10mm}\begin{tabular}{|l@{}|l@{}t@{}|l@{}t@{}|l@{}t@{}|}

@@

\end{tabular}}
